R package qshift: A functional programming approach to shifted probability distributions

### 2016-08-31 docs/article.Rmd
article

---
title: 'R package qtdshift: A functional programming approach to shifted probability distributions'
author: "Veronica"
date: "August 25, 2016"
output: pdf_document
---

@@

Shifted distributions are used daily in many statistical fields.   
A shifted distribution is a distribution with a constant added, $Y = X + a$, where $X$ and $Y$ are random variables and $a$ is a constant value.  
The aim of this paper is to describe an R package, `qdtshift` which generates shifted distribution models for any distribution chosen. The package was realized using a functional programming approach and inspires its computational method to Andrea Spanò article "R package qdist: A functional programming approach to truncated probability ditributions", which provides an alternative method for computing from truncated probability distribution using a functional programming approach.

@@

R does not provide a specific package for shifted distribution models, but it provides package `stats` which contains a wide set of functions for standard distribution models: probability, density, quantile and random number generator functions. All them share a stable naming convention both for the names of the functions and in the first argument of these functions. In particular:

@@

At first view, this approach seems easy and cool but thinking about it, we realize that we would need to write a different function for each probability distribution we would like to work with. Functional programming approach could be the solution, as Andrea Spanò demonstrates in his article (see "R package qdist: A functional programming approach to truncated probability ditributions" for more details).

`qtdshift` package includes four general functions which compute shifted distribution models for any distribution chosen:

@@

`qtdshift` package is available at [https://github.com/veronicagiro/qshift](https://github.com/veronicagiro/qshift).

@@

```{r qtdshift, message=FALSE}
require(qtdshift)
sdnorm <- dshift("norm") # density function
spnorm <- pshift("norm") # probability function
sqnorm <- qshift("norm") # quantile function
srnorm <- rshift("norm") # random generation function
```

Talking about the computational process, we can say that package `qtdshift` is based on two key concepts being part of the functional nature of the R programming language:

@@

### Examples

@@

pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=sdnorm_type)) +
  geom_line(size=1) +
  scale_color_manual(values = c("not_shifted" = "cyan3", 
                                "positive_shifted" = "#89FF99", 
                                "negative_shifted" = "#FF89B4")) +
  xlab("Value") + ylab("Density") +
  ggtitle("Density plot for shifted and non-shifted normal distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=12),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

@@

pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=spnorm_type)) +
  geom_line(size=1) +
  scale_color_manual(values = c("not_shifted" = "cyan3", 
                                "positive_shifted" = "#89FF99", 
                                "negative_shifted" = "#FF89B4")) +
  xlab("Quantile") + ylab("Probability") +
  ggtitle("Probability plot for shifted and non-shifted normal distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=12),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

@@

pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=sqnorm_type)) +
  geom_line(size=1) +
  scale_color_manual(values = c("not_shifted" = "cyan3", 
                                "positive_shifted" = "#89FF99", 
                                "negative_shifted" = "#FF89B4")) +
  ylab("Quantile") + xlab("Probability") +
  ggtitle("Quantile plot for shifted and non-shifted normal distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=12),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

@@

pl <- pl <- ggplot(data = df, mapping = aes(x=value, group=srnorm_type, fill=srnorm_type)) +
  geom_density(alpha=0.6) +
    scale_fill_manual(values = c("not_shifted" = "cyan3", 
                                "positive_shifted" = "#89FF99", 
                                "negative_shifted" = "#FF89B4")) +
  ggtitle("Density plot for shifted and non-shifted normal distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=12),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

@@

@@

```{r qtdshift_unif}
sdunif <- dshift("unif")
spunif <- pshift("unif")
squnif <- qshift("unif")
srunif <- rshift("unif")
```

Now the newly created functions are ready for computing from shifted probability models.  
We limited Uniform support between 6 and 12.

@@

pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=sdunif_type)) +
  geom_line(size=1) +
  scale_color_manual(values = c("not_shifted" = "cyan3", 
                                "positive_shifted" = "#89FF99", 
                                "negative_shifted" = "#FF89B4")) +
  xlab("Value") + ylab("Density") +
  ggtitle("Density plot for shifted and non-shifted uniform distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=12),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

@@

pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=spunif_type)) +
  geom_line(size=1) +
  scale_color_manual(values = c("not_shifted" = "cyan3", 
                                "positive_shifted" = "#89FF99", 
                                "negative_shifted" = "#FF89B4")) +
  xlab("Quantile") + ylab("Probability") +
  ggtitle("Probability plot for shifted and non-shifted uniform distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=12),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

@@

pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=squnif_type)) +
  geom_line(size=1) +
  scale_color_manual(values = c("not_shifted" = "cyan3",  
                                "positive_shifted" = "#89FF99", 
                                "negative_shifted" = "#FF89B4")) +
  ylab("Quantile") + xlab("Probability") +
  ggtitle("Quantile plot for shifted and non-shifted uniform distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=12),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

@@

pl <- pl <- ggplot(data = df, mapping = aes(x=value, group=srunif_type, fill=srunif_type)) +
  geom_density(alpha=0.6) +
    scale_color_manual(values = c("not_shifted" = "cyan3", 
                                "positive_shifted" = "#89FF99", 
                                "negative_shifted" = "#FF89B4")) +
  xlab("Value") + ylab("Density") +
  ggtitle("Density plot for shifted and non-shifted uniform distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=12),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

>>>>

---
title: 'R package qshift: A functional programming approach to shifted probability distributions'
author: "Veronica"
date: "August 25, 2016"
output:
  html_document:
    theme: null
    highlight: null
---

@@

Shifted distributions are used daily in many statistical fields.  
Given a random variable $X$ with distribution function $F(x)$, a shifted random variable $Y$ from $X$ is a random variable with a constant, $k$, added, defined as $Y = X+k$, with distribution function $F(x-k)$. $Y$ support changes according to shift value.

The aim of this paper is to describe an R package, `qshift` which generates shifted distribution functions for any distribution chosen. The package was realized using a functional programming approach and inspires its computational method to Andrea Spanò article "R package qdist: A functional programming approach to truncated probability ditributions", which provides an alternative method for computing from truncated probability distribution using a functional programming approach.

@@

R does not provide a specific package for shifted distribution functions, but it provides package `stats` which contains a wide set of functions for standard distribution functions: probability, density, quantile and random number generator functions. All them share a stable naming convention both for the names of the functions and in the first argument of these functions. In particular:

@@

At first view, this approach seems easy and cool but thinking about it, we realize that we would need to write a different function for each probability distribution we would like to work with. Functional programming approach could be a solution, as Andrea Spanò demonstrates in his article (see "R package qdist: A functional programming approach to truncated probability ditributions" for more details).

`qshift` package includes four general functions which compute shifted distribution models for any distribution chosen:

@@

`qshift` package is available at [https://github.com/veronicagiro/qshift](https://github.com/veronicagiro/qshift).

@@

```{r qshift, message=FALSE}
require(qshift)
sdnorm <- dshift("norm") # density function
spnorm <- pshift("norm") # probability function
sqnorm <- qshift("norm") # quantile function
srnorm <- rshift("norm") # random generation function
```

Talking about the computational process, package `qshift` is based on two key concepts being part of the functional nature of the R programming language:

@@

## Examples

@@

pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=sdnorm_type)) +
  geom_line(size=1) +
  scale_color_manual(values = c("not_shifted" = "cyan3", 
                                "positive_shifted" = "#89FF99", 
                                "negative_shifted" = "#FF89B4")) +
  xlab("Value") + ylab("Density") +
  ggtitle("Density plot for shifted and non-shifted normal distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=10),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

@@

pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=spnorm_type)) +
  geom_line(size=1) +
  scale_color_manual(values = c("not_shifted" = "cyan3", 
                                "positive_shifted" = "#89FF99", 
                                "negative_shifted" = "#FF89B4")) +
  xlab("Quantile") + ylab("Probability") +
  ggtitle("Probability plot for shifted and non-shifted normal distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=10),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

@@

pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=sqnorm_type)) +
  geom_line(size=1) +
  scale_color_manual(values = c("not_shifted" = "cyan3", 
                                "positive_shifted" = "#89FF99", 
                                "negative_shifted" = "#FF89B4")) +
  ylab("Quantile") + xlab("Probability") +
  ggtitle("Quantile plot for shifted and non-shifted normal distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=10),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

@@

pl <- pl <- ggplot(data = df, mapping = aes(x=value, group=srnorm_type, fill=srnorm_type)) +
  geom_density(alpha=0.6) +
    scale_fill_manual(values = c("not_shifted" = "cyan3", 
                                "positive_shifted" = "#89FF99", 
                                "negative_shifted" = "#FF89B4")) +
  ggtitle("Density plot for shifted and non-shifted normal distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=10),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

@@

As expected, the Normal shifted distribution is a Normal distribution with changed mean. The change on the mean is equal to the shift.

@@

```{r qshift_unif}
sdunif <- dshift("unif")
spunif <- pshift("unif")
squnif <- qshift("unif")
srunif <- rshift("unif")
```

We limited Uniform support between 6 and 12.

@@

pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=sdunif_type)) +
  geom_line(size=1, alpha =0.8) +
  scale_color_manual(values = c("not_shifted" = "cyan3", 
                                "positive_shifted" = "#89FF99", 
                                "negative_shifted" = "#FF89B4")) +
  xlab("Value") + ylab("Density") +
  ggtitle("Density plot for shifted and non-shifted uniform distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=10),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

@@

pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=spunif_type)) +
  geom_line(size=1, alpha =0.8) +
  scale_color_manual(values = c("not_shifted" = "cyan3", 
                                "positive_shifted" = "#89FF99", 
                                "negative_shifted" = "#FF89B4")) +
  xlab("Quantile") + ylab("Probability") +
  ggtitle("Probability plot for shifted and non-shifted uniform distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=10),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

@@

pl <- ggplot(data = df, mapping = aes(x=x, y=value, col=squnif_type)) +
  geom_line(size=1, alpha =0.8) +
  scale_color_manual(values = c("not_shifted" = "cyan3",  
                                "positive_shifted" = "#89FF99", 
                                "negative_shifted" = "#FF89B4")) +
  ylab("Quantile") + xlab("Probability") +
  ggtitle("Quantile plot for shifted and non-shifted uniform distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=10),
  plot.title = element_text(size=14), axis.title=element_text(size=12))

@@

pl <- pl <- ggplot(data = df, mapping = aes(x=value, group=srunif_type, fill=srunif_type)) +
  geom_density(alpha=0.6) +
    scale_color_manual(values = c("not_shifted" = "cyan3", 
                                "positive_shifted" = "#89FF99", 
                                "negative_shifted" = "#FF89B4")) +
  xlab("Value") + ylab("Density") +
  ggtitle("Density plot for shifted and non-shifted uniform distribution") +
  theme(legend.title=element_blank(), legend.text= element_text(size=10),
  plot.title = element_text(size=14), axis.title=element_text(size=12))
